Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-05-27 manuscript/311 manuscript.tex
Updated Tex file and accompanying png files

@@

New York City (NYC) has emerged as a forerunner in the open data
movement, marked by the enactment of the Open Data Law in 2012
\citep{zuiderwijk2014open}. This landmark legislation led to the
creation of the \href{https://opendata.cityofnewyork.us}{NYC Open Data
  portal}, which today hosts an impressive array of 2.700 datasets
from 80 different city agencies. This resource has become invaluable
for researchers in various fields as well an enabling a high degree of local government transparency

In health, datasets have enabled
significant studies on facets of health and healthcare delivery
\citep{cantor2018facets, shankar2021data}. In the realm of urban
development, data has been instrumental in advancing smart city
initiatives \citep{neves2020impacts}. Additionally, transportation
research has benefited greatly from this wealth of data, aiding in the
understanding of urban mobility and infrastructure
\citep{gerte2019understanding}.

NYC's Open Data initiative not only
exemplifies commitment to transparency and public engagement but also
illustrates how open data can be a powerful tool in addressing complex
urban challenges.

@@

In the realm of collaborative data management,
\citet{beheshti2019datasynapse} underscore the significance of
cooperative environments for managing and sharing social data
effectively. This aspect of data curation gains further relevance in
the research by \citet{mclure2014data}, which delves into the specific
practices and needs within data curation communities. The practical
implications of data curation are vividly illustrated in the context
of public health and global challenges. \citet{cantor2018facets}
demonstrate the utility of curated open data in evaluating community
health determinants. Furthermore, the COVID-19 pandemic serves as a
real-world example, with \citet{shankar2021data} observing the
critical role of collective data curation efforts in managing and
responding to the crisis.

Collectively, these studies not only highlight the multifaceted nature of data curation but also emphasize
its indispensable role in enhancing the applicability and value of
open data across various domains.

The contributions of this paper are twofold. First, we delve into
the specifics of data curation challenges using the NYC 311 Service
Request Data as a case study. This renowned and frequently viewed 
dataset serves as a prime example for examining key issues in data curation, 
including data validity, consistency, and curation efficiency. 
We illustrate these points with live examples drawn from our processing 
of the 311 data.

@@

The remainder of this paper is organized as follows. 
Section~\ref{sec:data} offers a brie review of the 
history and current status of the 311 requests data. In 
Section~\ref{sec:issues}, we identify and list potential issues 
impacting data quality and curation efficiency. 
Section~\ref{sec:improve} is dedicated to providing 
actionable suggestions for mitigating or resolving these identified issue. 
Following this, Section~\ref{sec:protocol} outlines a series of general 
principles for the release of open data, drawing from our findings. 
The paper concludes with a discussion in Section~\ref{sec:disc}, encapsulating 
the key insights and implications of our research.

\section{NYC 311 Service Requests Data} \label{sec:data}

The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. The evolution of this system can be traced from
its initial implementation as a simple inquiry channel to a
comprehensive data management system that handles millions of requests
annually. Key milestones in its development include:

\begin{itemize}
 	\item 2003 - NYC 311 system  goes live with phone-based call center only.
   	\item 2009 - 311 online \& mobile apps launched. Today 60\% of all Service Requests come from mobile \& online channels.
   	\item 2019 - Major software upgrade of 311 system with enhanced capabilities and cloud-based resilience. 	
      	\item 2020 - In August, driven by the COVID pandemic, monthly 311 Service Requests hit an all-time high of 348,463. Seven of the Top 10 busiest days occur in 2020.
      	\item 2021 - 311 System is expanded to into the MTA's city subway system, the largest expansion since its inception.
   	\item 2023 - Record high year for 311 with 3.23 million Service Requests.
\end{itemize}

Today, the NYC 311 data system is a robust platform that manages a
vast array of urban living-related inquiries. As of 2024, the system handles approximately 3.5 million service 
requests per year, covering categories such as noise complaints, illegal parking, heat/hot water complaints, abandoned vehicles,
and unsanitary conditions. The data, managed through a tailored application built on the Microsoft Dynamics platform, is a valuable resource for city
administration and policy-making. It is publicly accessible through the 
\href{https://opendata.cityofnewyork.us/}{NYC Open Data Portal}. The online Portal provides access
to all the data sets (approx. 2700) and provides a convenient method for querying, grouping, aggregating, geo-mapping, visualization, and exporting
results. This open data initiative enables not only governmental
transparency but also empowers researchers, civic developers, and the
general public. Examples of its use include:

\begin{itemize}
	\item NYC restaurant violations
	\item Popular baby names
	\item Mapping of car crashes involving pedestirans
	\item Mapping of sidewalk widths in NYC 
	\item Visualization of NYC High School and College enrollment
	\item Interactive dashboard to filter and view charges for jail inmates
	\item Internet access points.
	\item ...and 220 different complaint types. 
\end{itemize}

@@

Despite its success, the system faces many challenges such as:
\begin{itemize}
	\item Data timeliness, accuracy, and consistency
	\item Difficulties correlating data over long time periods, e.g. over a 10-year period
	\item Data anonymization and handling Personally identifiable information (PII)
	\item Integration with stand-alone systems at selected NYC Agencies
	\item Managing API usage, authentication, and load for 3rd party users
	\item Incorporating ongoing technology changes and upgrades 
	\item Understanding the role of chatbots and AI to provide online customer service
\end{itemize}

These (and other) challenges are constantly being addressed by the various agency open data managers
as well as the \href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology and Innovation (OTI)} which
provides the technology support for the open data system.  
Many of the underlying datasets are updated daily, usually running 1-2 days behind from creation.
However, a number of datasets are updates monthly (such as restaurant inspections), and some are
not updated for one or more years (such as community board boundaries). This poses problems when 
merging these dataset with one another, as would be expected. Typically, the data must be
exported and manipulated in another program

Data accuracy and consistency is typically established between OTI and the other contributing
City Agencies, such as NYPD, Housing Preservation and Development (HPD), Parks \& Recreation, etc.,
via Service Level Agreements (SLAs). However, enforcement of those SLAs is often challenging. The 
result is that often data accuracy and logical consistency is occasionally compromised.

The NYC 311 organization, part of OTI, operates a highly tailored and sophisticated case
management system based, as noted, on Microsoft Dynamics. However, when the original 311
system was developed in 2003, a decision was made to allow select NYC Agencies to
maintain their own local computer systems, such as the mainframe systems at the NYC Department
of Sanitation (DSNY) and others. And when the 311 system was completely upgraded in 2019, those
same integration were `honored` in order to reduce the scope of the project. (Perhaps in
retrospect, while a good budgetary decision, perhaps not such a good technology decision.)

The result was that at least nine NYC Agencies have local systems that are integrated either via
an API or custom code. These Agencies include:

@@

One result of these integrations, and other Agencies that use the 311 APIs, is that some of the data accuracies
and data consistency checks inherent in the 311 system are not applied uniformly, and in some cases not at all.
Errors often creep in from other Agency systems and are not easily corrected in the 311 application itself,
but rather require correction at the source Agency.

One area that will almost certainly remain an ongoing challenge is the incorporation of new software, hardware,
and cloud services which are constantly emerging.  Telephonic integration recently was reconfigured due to a new
telephone vendor. Desktop software at the 311 Agency are subject to the same upgrades that home users feel. And while the 
311 system was designed to be easily configured with new features, some emerging business issues as well as
technologies simply require additional software engineering efforts to accommodate.

One area in particular which is heavily influencing customer outreach for the 311 system is the use of
online chatbots. Increasing these are being influenced by Artificial Intelligence (AI) advances. It is a given that
a number of citizen's interactions with City government could be handled accurately and quickly by AI-driven
applications. How to do that while preserving customer privacy and anonymization are challenging issues.

@@

\begin{itemize}
	\item Advice on shelters during hurricanes and winter blizzards
	\item Countless inquires during the COVID pandemic (testing centers, vaccine sites, mobile vaccine clinics, etc.)
	\item Enforcement of standards between landlords and tenants.
	\item Re-allocation of NYC's yellow taxi routes based upon an insightful analysis conducted by the Taxi \& Limousine Commission (TLC)
	\item Improved responsiveness for City Agencies with direct client-facing responsibilities such as Parks \& Rec, Dept of Transportation, DSNY, etc.
	\item Increased 311 SRs related to Homelessness resulted in the launch of the City's Homeless Outreach \& Mobile Engagement Street Action Teams (HOME-STAT)
	which includes street canvassers and increases in homeless housing facilities.
	\item A significant increase in graffiti complains in 2017 led directly to the `Graffiti-Free NYC'' campaigns and increased funding for removal of graffiti in public spaces
	\item Enhanced geospatial data regarding the location of incidents, the responding Agency, and actions taken.
	\item Establishment of Agency SLAs in regard to certain complaint types, such as residential noise, homelessness person assistance, street light outages, 
	rat sightings, and other neighborhood quality of life issues.
\end{itemize}

The NYC 311 service exemplifies the dynamic nature of urban data
management and its critical role in modern governance. It provides an
invaluable resource for data scientists to practice data science
activities, including data curation.

We extracted  311 Service Request (SR) data from the period of CY 2022-2023, a dataset consisting of
6.4 million records. This dataset serves as the primary source for the analysis effort. This is a large dataset which enables 
observation of some rarely seen events, data inaccuracies, and logical inconsistencies. It provides enough data to examine trends and data outlier analysis. 
We query the data directly from the NYC Open Data Portal and then export the data in CSV format for analysis in a custom written R programs.

Additionally, we extracted two longer-term datasets, a 5-year look (2019-2023) and a 10-year look (2014-2023) ; some 14  and 27 million rows respectively, albeit with just a few columns. 
These larger datasets afford an opportunity for long-term trend observations, confirming the analysis using the 2-year dataset. The 2022-2023 dataset
was used exclusively for the detailed, field-level analysis that comprises the majority of this effort.

\section{10-year Analysis \& Trends} \label{sec:history}

Since the NYC 311 System went live in 2003, the usage has grown at an annual
rate of 4.6\%. For this study, we also looked at a 10-year period (2014-2023) to
observe certain system performance characteristics over a longer time-frame . 
During that time-frame, the number of 311 SRs grew 50\% as shown in the chart below which shows the yearly SR counts.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{Yearly.png}
  \caption{SR counts by year for a 10-year period}
  \label{fig:yearly-counts}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{Monthly.png}
  \caption{SR counts by month for a 10-year period}
  \label{fig:monthly-counts}
\end{figure}

A quick look at the corresponding  daily tend over that same 10-year period is also interesting, as it shows both significant spikes
and significant dips; a noisy graph. While the spikes on 2020-08-04 (24,415 SRs) and 2020-08-05 (19,560 SRs) are COVID related, we are unable to 
ascertain any reason for the significant spike on Saturday, 2023-09-29 where 17,962 SRs were created.

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{Daily.png}
  \caption{SR counts by day for a 10-year period}
  \label{fig:daily-counts}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{Calendar-Month.png}
  \caption{SR counts by calendar month for a 10-year period}
  \label{fig:calendar-months-counts}
\end{figure}

Of note, the lowest SR counts-per-day occur on Christmas  period (+/- a day), in 5 of the Bottom 10 SR counts-per-day. The summer moths and early
fall months make up all of the Top 10 SR counts-per-day. As previously noted, the all-time daily high for SRs occurred on Tuesday, 2020-08-04 with the 10-year
low number occurs on Christmas Day in 2016, a Sunday. Note also that the difference between the peak day (24,415) and the minimum day (2965) varies
by a factor of 8X, and the aveage daily count of 7489 -- all of which creates a challenging  system capacity design issue.

\begin{table}[htbp]
	\begin{minipage}[t]{0.35\textwidth} % Adjust the width of the minipage as needed
	\centering
	\small % Adjust the font size
	\caption{Top 10 days}
	\label{tab:top_ten_counts}
		\begin{tabular}{@{}lll@{}}
			\toprule
			Date & Count & Rank \\
			\midrule
			2020-08-04 & 24415 & 1 \\
			2020-08-05 & 19560 & 2 \\
			2023-09-29 & 17962 & 3 \\
			2020-07-05 & 16916 & 4 \\
			2020-06-21 & 15883 & 5 \\
			2020-06-20 & 15825 & 6 \\
			2020-07-04 & 15794 & 7 \\
			2021-09-02 & 15205 & 8 \\
			2020-06-28 & 14057 & 9 \\
			2021-10-28 & 13575 & 10 \\
			\bottomrule
		\end{tabular}
	\end{minipage}\hfill % Adjust the space between the tables
	\begin{minipage}[t]{0.35\textwidth} % Adjust the width of the minipage as needed
	\centering
	\small % Adjust the font size
	\caption{Bottom 10 days}
	\label{tab:bottom_ten_counts}
		\begin{tabular}{@{}lll@{}}
			\toprule
			Date & Count & Rank \\
			\midrule
			2014-12-25 & 2965 & 1 \\
			2015-12-25 & 3247 & 2 \\
			2014-04-20 & 3287 & 3 \\
			2015-12-26 & 3405 & 4 \\
			2016-11-06 & 3419 & 5 \\
			2014-07-04 & 3455 & 6 \\
			2014-12-27 & 3534 & 7 \\
			2014-09-14 & 3563 & 8 \\
			2014-02-02 & 3565 & 9 \\
			2016-12-25 & 3587 & 10 \\
			\bottomrule\\
		\end{tabular}
	\end{minipage}
\end{table}

And although the highest number of SRs occur in July and the lowest in February,
when the \#-of-SRs-per-day metric is computed by adjusting for the differing number of days for each month -- the
busiest month is shown to be June (followed by July and September). And similarly the quietest months are April, March, and 
February. In general, the winter months see fewer SRs while the summer months show 
an increase in activity, as previously noted.

\begin{table}[ht]
    \centering
    \small
    \begin{tabular}{lS[table-format=7.0]S[table-format=6.0]} % 'l' for left-aligned, 'S' for siunitx number alignment
        \toprule
        \textbf{Month}     & \textbf{Count}   & \textbf{Count per Day} \\ 
        \midrule
        June  & 2454156 & 81805 \\ 
        July  & 2461525 & 79404 \\ 
        September & 2352822 & 78427 \\ 
        October & 2407238 & 77653 \\ 
        August & 2401731 & 77475 \\ 
        May  & 2334466 & 75305 \\ 
        November  & 2236037 & 74535 \\ 
        January   & 2271524 & 73275 \\ 
        December  & 2199765 & 70960 \\ 
        February  & 1996408 & 70795 \\ 
        March & 2174349 & 70140 \\ 
        April  & 2058169 & 68606 \\ 
       \bottomrule
    \end{tabular}
    \caption{Monthly counts sorted by counts-per-day}
    \label{tab:monthly_counts}
\end{table}

@@

Two areas that are important in this data cleansing analysis effort are the responsible NYC Agency, and the actual type of complaint.
Given both the size of the City of New York government, it's typically necessary to identify the responsible Agency as well as the type of complaint in order to 
clarify the responsible area and assist in troubleshooting any discrepancy.

In discovering a data error, we typically observe two trends; either a single (or a small few) Agencies are responsible (such as Dept of Transportation), or it is a system-wide issue that foolows the general distribution
of  Service Requests (SRs) across the system. Let's take a brief look at how this plays out across the system, so when we identify a
issue, it will be useful to see how that issue flows back to a specific Agency, and thus establish responsibility.

Although the 311 SRs involve 16 different agencies, there are the ``big six'' Agencies that comprise
90\% of the complaints. These six are (in order):

@@

\begin{figure}[htbp]
  \centering
	  \includegraphics[width=\textwidth]{SRs_by_Agency.png}
	  \caption{SR counts by Agency and Cumulative Percentage}
	  \label{fig:SR_counts_by_Agency}
\end{figure}

It is also interesting to note the distribution of the type of complaints in the 311 SR dataset; it is heavily skewed to  a select few complaints.

\begin{figure}[htbp]
  \centering
	  \includegraphics[width=\textwidth]{SRs_by_Complaint_Type.png}
	  \caption{Top 20 SR Complaints and Cumulative Percentage}
	  \label{fig:SR_complaints}
\end{figure}

The top 20 complaints contain several of the eight different types of ``Noise'' complaints (Residential, Commercial, Street, Helicopter, etc.). These Noise-related
complaints total 1.435 million in the 2022-2023 dataset, a full 22\% of all complaint types, which is the most frequently occurring type. (Note: Ilegal Parking is 
second, followed by Heat/Hot Water.) Noise complaints are also handled by NYPD, as is Illegal Parking and Blocked Driveway complaints, which mostly
accounts for the top ranking of 311 SRs being routed to the NYPD.

There are also some rather curious and unusual Service Requests in the 311 data: 
tanning, tattooing, trans fat, unsanitary pigeon condition, illegal animal kept as pet, harboring bees/wasps, and radioactive material. 
New Yorkers certainly live interesting lives.

>>>>

\hyphenpenalty=1000

@@

New York City (NYC) has emerged as a forerunner in the open data
movement, marked by the enactment of the Open Data Law in 2012
\citep{zuiderwijk2014open}. This landmark legislation led to the
creation of the \href{https://opendata.cityofnewyork.us}{NYC Open Data
  portal}, which today hosts an impressive array of 2700 datasets
from 80 different city agencies. This resource has become invaluable
for researchers in various fields as well an enabling local government transparency

In health, datasets have enabled
significant studies on facets of health and healthcare delivery
\citep{cantor2018facets, shankar2021data}. In the realm of urban
development, data has been instrumental in advancing smart city
initiatives \citep{neves2020impacts}. Additionally, transportation
research has benefited greatly from this wealth of data, aiding in the
understanding of urban mobility and infrastructure
\citep{gerte2019understanding}. NYC's Open Data initiative not only
exemplifies commitment to transparency and public engagement but also
illustrates how open data can be a powerful tool in addressing complex
urban challenges. Examples of some of the more popular NYC datasets include:

\begin{itemize}
	\item NYC restaurant violations
	\item Popular baby names
	\item Mapping of car crashes involving pedestrians
	\item Mapping of sidewalk widths in NYC 
	\item Visualization of NYC High School and College enrollment
	\item An interactive dashboard to filter and view charges for jail inmates
	\item Location of City-wide free Internet access points
	\item ...and 220 other complaint types. 
\end{itemize}

@@

In the realm of collaborative data management,
\citet{beheshti2019datasynapse} underscore the significance of
cooperative environments for managing and sharing social data
effectively. This aspect of data curation gains further relevance in
the research by \citet{mclure2014data}, which delves into the specific
practices and needs within data curation communities. The practical
implications of data curation are vividly illustrated in the context
of public health and global challenges. \citet{cantor2018facets}
demonstrate the utility of curated open data in evaluating community
health determinants. Furthermore, the COVID-19 pandemic serves as a
real-world example, with \citet{shankar2021data} observing the
critical role of collective data curation efforts in managing and
responding to the crisis. Collectively, these studies not only highlight the 
multifaceted nature of data curation but also emphasize
its indispensable role in enhancing the applicability and value of
open data across various domains.

The contributions of this paper are twofold. First, we delve into
the specifics of data curation challenges using the NYC 311 Service
Request Data as a case study. This renowned and frequently viewed 
dataset serves as a prime example for examining key issues in data curation, 
including data validity, consistency, and curation efficiency. 
We illustrate these points with live examples drawn from our processing of the 311 SR data.

@@

The paper is organized as follows:
Section~\ref{sec:history} offers a brief review of the history of the 311 system. In 
Section~\ref{sec:trends} we take a look at long-term trends presented via a 10-year analysis.
Section~\ref{sec:issues} offers a general discussion of data cleansing issues
impacting data quality and curation efficiency. Section~\ref{sec:Structural} examines
the technical dimensions of the dataset, the structural issues. In Section~\ref{sec:datatypes} we examine the data fields
for compliance with the stated data type as found in the Data Dictionary. Section~\ref{sec:blanks} looks
at the data fields as regards missing, blank, or N/A entries. Section~\ref{sec:domain} explores
how data fields comply with a domain of legal or acceptable values. Section~\ref{sec:inconsistencies}
deals with the important issues surrounding logical inconsistencies and concerning patterns in the data. Section~\ref{sec:precision} 
explores the age-old issue of precision versus accuracy. Section~\ref{sec:duplicates}
identifies duplicate and redundant data fields. Section~\ref{sec:improvements} provides 
actionable suggestions for mitigating or resolving identified issues. Following this, Section~\ref{sec:protocol} outlines a series of general 
principles for the release of open data, drawing from our findings. The paper concludes with a discussion 
in Section~\ref{sec:discussion}, encapsulating the key insights and implications of our research.

A note on naming conventions: dataset field names are typically named using ``snake\_case'' where
each word in the variable name is written in lowercase, and words are separated by underscores,
e.g created\_date, complaint\_type, zip\_code, etc..

\section{NYC 311 System History} \label{sec:history}

The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. The evolution of this system can be traced from
its initial implementation as a simple inquiry channel via telephone to a
comprehensive data management system that handles millions of requests
annually. Each inquiry is logged as a complaint in the 311 system.
Key milestones in the 311 system development include:

\begin{itemize}
 	\item 2003 - NYC 311 system goes live with a phone-based call center only.
   	\item 2009 - 311 online \& mobile apps launched. Today 60\% of all Service Requests come from mobile \& online channels.
   	\item 2019 - Major software upgrade of 311 system with enhanced capabilities, dynamic load-balancing, and cloud-based resilience. 	
      \item 2020 - In August 2020, driven by the COVID pandemic, monthly 311 Service Requests hit an all-time high of 348,463. Seven of the Top 10 busiest days of all time occur in 2020.
      \item 2021 - 311 System is expanded to into the MTA's city subway system, the largest expansion since its inception.
   	\item 2023 - Record high year for 311 with 3.23 million Service Requests.
\end{itemize}

Today, the NYC 311 data system is a robust platform that manages a
vast array of urban living-related inquiries. The system handles over 3 million service 
requests (SRs) per year, encompassing such complaints as street noise, illegal parking, heat/hot water, abandoned vehicles,
and unsanitary conditions. The data, managed through a tailored application built on the Microsoft Dynamics and Azure platform, is a valuable resource for city
administration and policy-making. The 311 data is publicly accessible through the 
\href{https://opendata.cityofnewyork.us/}{NYC Open Data Portal} which provides access
to all the data sets (approx. 2700) and provides a convenient method for querying, grouping, aggregating, geo-mapping, visualization, and exporting
results. This open data initiative enables not only governmental transparency but also empowers 
researchers, civic developers, and the general public.

@@

Despite its success, the system constantly faces challenges such as:

\begin{itemize}
	\item Data timeliness, accuracy, and consistency
	\item Difficulties correlating data over long time periods, e.g. over a 10-year period, e.g. Agency name changes, new complaint types.
	\item Data anonymization and handling of Personally identifiable information (PII)
	\item Integration with stand-alone systems at selected NYC Agencies
	\item Managing API usage, authentication, and load for 3\textsuperscript{rd} party users, of which there are many
	\item Incorporating ongoing technology changes and upgrades 
	\item Understanding the role of chatbots and AI to provide online customer service, along with other emergent technologies
\end{itemize}

These and other challenges are continually addressed by the various Agency open data managers
as well as the \href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology and Innovation (OTI)} which
provides the technology support for the open data system.  
Many of the underlying datasets are updated daily, usually running 1-2 days behind from creation.
However, a number of datasets are updated monthly (such as restaurant inspections), and some are
not updated for one or more years (such as community board boundaries). This poses problems when 
merging dataset with others, as would be expected. Typically, the data must be
exported and manipulated in another, external program. The authors found this to be the case with 
the large 311 dataset employed.

Data accuracy and consistency is typically established between OTI and the other contributing
City Agencies, such as NYPD, Housing Preservation and Development (HPD), Parks \& Recreation, etc.,
via Service Level Agreements (SLAs). However, enforcement of those SLAs is often challenging. The 
result is that often data accuracy and logical consistency is occasionally compromised. However, the Open Data
Team is quite responsive to inquires from users. The authors have provided many such 
suggestions during the course of this research effort, and the most current data shows improvements
in many of the areas highlighted. Questions and suggestion  can be provide here: 
\href{https://opendata.cityofnewyork.us/engage/}{Contact Us}.

The NYC 311 organization, part of OTI, operates a highly tailored and sophisticated case
management. However, one constraint was particularly challenging during the recent complete overhaul
of the software and hardware in 2019. When the original 311system was developed in 2003, 
a decision was made to allow select NYC Agencies to maintain their own local computer systems, 
such as the mainframe systems at the NYC Department of Sanitation (DSNY) and others. 
And when the 311 system was completely upgraded in 2019, those
same integrations were honored in order to reduce the scope of the project. Perhaps in
retrospect, while a good budgetary and policy decision, it was perhaps not such a good technology decision.

At least nine NYC Agencies have local systems that are integrated directly to the 311 system either via
an API or custom code. These include:

@@

One unfortunate side-effect from these integrations, and other Agencies that use the 311 APIs, is that some of the data consistency
checks which are inherent in the 311 system are not applied uniformly, and in some cases not at all, to the integrated data streams.
Errors often creep in from other Agency systems and are not easily corrected or detected in the 311 application itself,
but rather require correction at the source Agency; challenging.

One area that will almost certainly remain an ongoing issue is the ongoing incorporation of new software, hardware,
and cloud services which are constantly changing while new approaches emerge. The 311 telephonic integration recently was reconfigured due to a new
telephonic vendor. Desktop software at the 311 Agency are subject to the same upgrades that home and business users experience. And while the 
311 system was designed to be easily configured and updated with new features, some emerging business issues as well as
technologies nonetheless require additional software engineering efforts to accommodate, along with the concomitant budgets.

One area in particular which is heavily influencing customer outreach for the 311 system is the appropriate use of
online chatbots and assistants. Increasing these features are influenced by artificial intelligence (AI) advances. It is a given that
a number of citizen's interactions with City government could be handled accurately and quickly by AI-driven
applications thus improving timeliness and accuracy. But how to do that while preserving customer privacy and anonymization
and user engagement are challenging issues.

@@

\begin{itemize}
	\item Providing advice on shelters during hurricanes, heat emergencies, and winter blizzards
	\item Handling countless inquires during the COVID pandemic (testing centers, vaccine sites, mobile vaccine clinics, etc.)
	\item Enforcement of standards between landlords and tenants
	\item Re-allocation of NYC's yellow taxi routes based upon an insightful analysis conducted by the Taxi \& Limousine Commission (TLC)
	\item Improved responsiveness for City Agencies with regards to direct client-facing responsibilities such as Parks \& Rec, Dept of Transportation, DSNY, etc.
	\item Increased 311 SRs related to Homelessness resulted in the launch of the City's Homeless Outreach \& Mobile Engagement Street Action Teams (HOME-STAT)
	which includes street canvassers and increases in homeless housing facilities.
	\item A significant increase in graffiti complains in 2017 led directly to the `Graffiti-Free NYC'' campaigns and increased funding for removal of graffiti in public spaces.
	\item Enhanced geospatial data regarding the location of incidents, the responding Agency, and actions taken across a myriad of complaint\_types
	\item Establishment of Agency SLAs in regard to certain complaint types, such as residential noise, homelessness person assistance, street light outages, 
	rat sightings, and other neighborhood quality of life issues.
\end{itemize}

The NYC 311 service exemplifies the dynamic nature of urban data
management and its critical role in modern governance. In addition it provides an
invaluable resource for data scientists to practice data science activities, including data curation.

We extracted 311 SR data for the period of CY 2022-2023, a dataset consisting of
6.4 million records. That dataset serves as the primary source for the analysis effort. This is a large dataset which enables 
observation of some rarely seen events, data inaccuracies, and logical inconsistencies. It provides enough data to examine trends and data outlier analysis. 
We queried the data directly from the NYC Open Data Portal and then exported the data in CSV format for analysis using custom written R programs.

Additionally, we extracted two longer-term datasets, a 5-year look (2019-2023) and a 10-year look (2014-2023) ; some 14 and 27 million rows respectively. 
These larger datasets afford an opportunity for long-term trend observations, confirming and supporting the analysis using the 2-year dataset. The 2022-2023 dataset
was used exclusively for the detailed, field-by-field level analysis which comprises the majority of this effort.

\section{10-year Analysis \& Trends} \label{sec:trends}

Since the NYC 311 System went live in 2003, the usage has grown at an annual
rate of 4.6\%. For this study, we also looked at a 10-year period (2014-2023) to
observe certain system performance characteristics over a longer time-frame . 
During that time-frame, the number of 311 SRs grew 50\% as shown in the chart below showing yearly SR counts.

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{Yearly.png}
  \caption{Yearly SR counts over 10-year period}
  \label{fig:yearly-counts}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{Monthly.png}
  \caption{Monthly SR counts over 10-year period}
  \label{fig:monthly-counts}
\end{figure}

A quick look at the corresponding  daily tend over a 5-year period (2019-2023) shows both significant spikes
and significant dips; a noisy graph. While the spikes on 2020-08-04 (24,415 SRs) and 2020-08-05 (19,560 SRs) are certainly COVID related, we were unable to 
ascertain any reason for the significant spike on Saturday, 2023-09-29 when 7,962 SRs were created.

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{Daily.png}
  \caption{Daily SR counts over 5-year period}
  \label{fig:daily-counts}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{Calendar-Month.png}
  \caption{Calendar Month SR counts over a 10-year period}
  \label{fig:calendar-months-counts}
\end{figure}

Of note, the lowest SR counts-per-day occur during the Christmas  holiday period encompassing fully 5 of the Bottom 10 SR counts-per-day. The summer months and early
fall months make up all of the Bottom 10 SR counts-per-day. As noted, the all-time daily high for SRs occurred on Tuesday, 2020-08-04 while the 10-year
low count occurs on Christmas Day in 2016, a Sunday. Note also that the difference between the peak day (24,415) and the minimum day (2965) varies
by a factor of 8X, and well exceeds the average daily count of 7489; all of which creates a challenging  system capacity and performance issue.

\begin{table}[H]
    \centering
    \small
    \begin{tabular}{@{}lll|lll@{}}
        \toprule
        \multicolumn{3}{c|}{Top 10 Days} & \multicolumn{3}{c}{Bottom 10 Days} \\
        \cmidrule(r){1-3} \cmidrule(l){4-6}
        Date & Count & Rank & Date & Count & Rank \\
        \midrule
        2020-08-04 & 24415 & 1 & 2014-12-25 & 2965 & 1 \\
        2020-08-05 & 19560 & 2 & 2015-12-25 & 3247 & 2 \\
        2023-09-29 & 17962 & 3 & 2014-04-20 & 3287 & 3 \\
        2020-07-05 & 16916 & 4 & 2015-12-26 & 3405 & 4 \\
        2020-06-21 & 15883 & 5 & 2016-11-06 & 3419 & 5 \\
        2020-06-20 & 15825 & 6 & 2014-07-04 & 3455 & 6 \\
        2020-07-04 & 15794 & 7 & 2014-12-27 & 3534 & 7 \\
        2021-09-02 & 15205 & 8 & 2014-09-14 & 3563 & 8 \\
        2020-06-28 & 14057 & 9 & 2014-02-02 & 3565 & 9 \\
        2021-10-28 & 13575 & 10 & 2016-12-25 & 3587 & 10 \\
        \bottomrule
    \end{tabular}
    \caption{Top 10 and Bottom 10 Days}
    \label{tab:combined_counts}
\end{table}

And although the highest number of SRs occur in July and the lowest in February,
when the \#-of-SRs-per-day metric is computed by adjusting for the differing number of days for each month -- the
busiest month is shown to be June (followed by July and September). Similarly the quietest months are April, March, and 
February. In general, the winter months see fewer SRs while the summer months show 
an increase in activity.

\begin{table}[H]
    \centering
    \small
    \begin{tabular}{lS[table-format=7.0]S[table-format=6.0]} % 'l' for left-aligned, 'S' for siunitx number alignment
        \toprule
        \textbf{Month}     & \textbf{Count}   & \textbf{Count per Day} \\ 
        \midrule
        June  & 2454156 & 81805 \\ 
        July  & 2461525 & 79404 \\ 
        September & 2352822 & 78427 \\ 
        October & 2407238 & 77653 \\ 
        August & 2401731 & 77475 \\ 
        May  & 2334466 & 75305 \\ 
        November  & 2236037 & 74535 \\ 
        January   & 2271524 & 73275 \\ 
        December  & 2199765 & 70960 \\ 
        February  & 1996408 & 70795 \\ 
        March & 2174349 & 70140 \\ 
        April  & 2058169 & 68606 \\ 
       \bottomrule
    \end{tabular}
    \caption{Monthly counts ordered by count-per-day}
    \label{tab:monthly_counts}
\end{table}

@@

Two important fields in this analysis effort are the responsible NYC Agency, and the type of complaint.
Given both the size of the City of New York government, it's necessary to identify the responsible Agency as well as the type of complaint in order to 
clarify the responsible party and assist in troubleshooting discrepancies.

When we discovered a data error, we typically observe two trends; either a single (or a few) Agencies are responsible (such as Dept of Transportation), or it is a system-wide issue that follows the general distribution
of  Service Requests (SRs) across the system. Thus we can often identify if it is an Agency specific problem, or a systemic effect affecting all Agencies.

Although the 311 SRs involve 16 different agencies, there are the ``big six'' Agencies that comprise
90\% of the complaints. These six are (in order of \% of complaints):

@@

\begin{figure}[H]
  \centering
	  \includegraphics[width=\textwidth]{SRs\_by\_Agency.png}
	  \caption{SR counts by Agency and Cumulative Percentage}

\end{figure}

Note the distribution of the complaint\_type; it is heavily skewed to a few select complaints with an exceptionally long tail.
This graph shows the top 20 complaint\_type(s) out of a total of 220.

\begin{figure}[H]
  \centering
	  \includegraphics[width=\textwidth]{SRs_by_Complaint_Type.png}
	  \caption{Top 20 SR complaint\_type and Cumulative Percentage}
	  \label{fig:SR_complaints}
\end{figure}

The top 20 complaints contain several of the eight different types of ``Noise'' complaints (Residential, Commercial, Street, Helicopter, etc.). These Noise-related
complaints total 1.435 million in the 2022-2023 dataset, a full 22\% of all complaint types, which is the most frequently occurring type. (Note: Ilegal Parking is second, followed by Heat/Hot Water.) Noise complaints are also handled by NYPD, as is Illegal Parking and Blocked Driveway complaints, which contributes
significantly to the top ranking of the NYPD as the responsible Agency.
